\documentclass[11pt]{article}

\include{header}

\begin{document}
\homework{10}{9 October}

\section{Sets}

In class we showed how to model a data-type of sets using lists and instantiating the type as an instance of the {\tt{Eq}} and {\tt{Show}} type classes.  Here's the code we developed in class with some additional functionality added.

\begin{verbatim}
module Set where

data Set a = Set [a] 

subset (Set xs) (Set ys) = null (filter (notin ys) xs)
  where notin xs x = not (x `Prelude.elem` xs)

instance  (Eq a ) => Eq (Set a) where
   xs == ys =  xs `subset` ys && ys `subset` xs

instance (Eq a, Show a) => Show (Set a) where
  show (Set xs) = fix $ show  $unique xs
    where fix xs = '{' : (take (length xs -2) (drop 1 xs) ++ "}")
          unique [] = []
          unique (x:xs) = x: unique (filter (/=x) xs)

empty = Set []

insert x (Set xs) = Set (x:xs)

elem x (Set xs) = x `Prelude.elem` xs

union (Set xs) (Set ys) = Set (xs ++ ys)

remove x (Set []) = (Set [])
remove x (Set (y:ys)) = if x == y then (Set ys) else insert y (remove x (Set ys))
\end{verbatim}

Recall that for sets, order of the elements and the multiplicity (how many
times they occur in the list) are not significant. This is reflected in set
equality - two sets are equal iff they contain the same elements.  A {\em{Bag}}
or {\em{Multiset}} is a structure there the order of the elements is not
significant but their multiplicities are.  We can characterize equality for
multisets if we have a function that counts the number of times an element
occurs in the mutiset.  For $m,n::Bag \; a$ and a function $count :: a
\rightarrow (Bag\,a) \rightarrow \,Int$ that counts the number of ioccurences
of an element in a bag, we define equality as follows:
\[\  m == n \definedAs \forall {}x::a .\; count \,x\, m == count\, x\, n\]
Note that for bags $m$ and $n$, $m$ is a subbag of $n$ if the following holds:
\[\  m \subseteq n \definedAs \forall {}x::a .\; count \,x\, m \le count\, x\, n\]
For bags $m$ and $n$ you could prove the following theorem holds:
\[\ m == n \Leftrightarrow m \subseteq n \wedge n\subseteq m\] This definition
provides an alternative path to implementing equality for bags.
\begin{exercise}
Use the code above as a template to implement a data-type of bags. You should
instantiate the type as an instace of the {\tt{Eq}} and {\tt{Show}} type
classes.  For the {\tt{show}} function I like a display the looks like ``$\mid$[
1,2,3]$\mid$'' for ``Bag [1,2,3]'' - but maybe you have a better idea.

You should also implement the following functions and constants:
\begin{verbatim}
subbag :: (Eq a) => Bag a -> Bag a -> Bool
empty :: Bag a
insert :: a -> Bag a -> Bag a
elem :: (Eq a) => a -> Bag a -> Bool
multiplicity :: (Num t, Eq a) => a -> Bag a -> t
union :: Bag t -> Bag t -> Bag t
remove :: (Eq a) => a -> Bag a -> Bag a
\end{verbatim}
\end{exercise}





\end{document}
% Local Variables:
% mode:latex
% comment-column:0
% comment-start: "% "
% compile-command: "pdflatex hw10d"
% fill-column:79
% End:

