\documentclass[11pt]{article}

\include{header}

\begin{document}
\homework{11}{4 October}

\section{Types and Terms}

Any discussion of type inference for a programming language involves
two languages: a language of type expressions and the programming
language itself.

We start with a simple programming language -- the $\lambda$-calculus with
pairs.  Let \[{\cal{V}}=\{x,y,z,w,x_1,y_1,z_1,w_1,\cdots\}\] be an unbounded
set of variables. The, the following grammar presents the language of
$\lambda$-terms.

\[\begin{array}{l}
\Lambda ::= x \mid (M\,N) \mid \lambda{}x.M \mid (M,N) \mid {\mathit{fst}}\;M \mid {\mathit{snd}} \, M \\
\;\;{\mathrm{where\ }} x\in{}{\cal{V}} {\mathrm{\ is\ a \ variable.}}\\
\;\;\;\;\;\;\;\;\;\;\;\;M,N\in\Lambda \; {\mathrm{\ are\  previously\  constructed\ \lambda-terms.}}\\
\;\;\;\;\;\;\;\;\;\;\;\;{\mathit{fst,snd}}\;{\mathrm{\ are\ constant\ symbols.}}
\end{array}
\]
The term $(M\,N)$ is an {\em{application}} (of $M$ to $N$.) The term
$\lambda{}x.M$ is an {\em{abstraction}} and is how functions are defined.  The
term $(M,\,N)$ is a pair and {\it{fst}} and {\it{snd}} are the projections
functions for pairs.  Eventually we will discuss computation in the language,
but for now, we are interested in determining if a term is well typed or not.

We represent the $\lambda$-terms in Haskell as follows:

\begin{smallprogram*}
\>data  Term =   Var String $\mid$ Ap Term Term $\mid$ Abs String Term $\mid$ Pair Term Term $\mid$ Fst Term $\mid$ Snd Term \\
\>        deriving (Eq,Show) \\
\end{smallprogram*}


Let ${\cal{VT}} = \{\alpha,\beta,\gamma,\alpha_1\cdots\}$ be an unbounded set
of type variables. The language of types is give by the following grammar.

\[\begin{array}{l}
T ::= \alpha \mid \tau_1 \rightarrow \tau_2 \mid \tau_1 \times \tau_2  \\
\;\;{\mathrm{where\ }} \alpha\in{}{\cal{VT}} {\mathrm{\ is\ a \ type\ variable.}}\\
\;\;\;\;\;\;\;\;\;\;\;\;\tau_1,\tau_2\in{}T \; {\mathrm{\ are\  previously\  constructed\ type\ expressions.}}\\
\end{array}\]

The type $\tau_1\rightarrow\tau_2$ denotes the type of functions from type
$\tau_1$ to $\tau_2$.  The type $\tau_1\times\tau_2$ denoted the type of pairs
where the first element is of type $\tau_1$ and the second is of type $\tau_2$.

The Haskell representation of type expressions can be given as follows:
\begin{smallprogram*}
\> data Op = Arrow $\mid$ Product     deriving (Eq,Show) \\
\> data  Type =   TVar  String $\mid$  BinType Op  Type Type     deriving Eq \\
\end{smallprogram*}

In this representation the type $\alpha\rightarrow(\beta\times\gamma)$ would
have the representation
\begin{smallprogram*}
\>  BinType Arrow (TVar "a") (BinType Product (TVar "b") (TVar "c"))
\end{smallprogram*}

Note that type variables occur as the leafs in the syntax trees of type
expressions.


We choose to instantiate the Haskell type {\it{Type}} as an instance of the
show type class as follows:

\begin{smallprogram*}
\> instance Show Type where \\
\>    show (TVar x) = x \\
\>    show (BinType Arrow t1 t2) = "(" ++ show t1 ++ " -> " ++ show t2 ++ ")" \\
\>    show (BinType Product t1 t2) = "(" ++ show t1 ++ " X " ++ show t2 ++ ")" \\
\end{smallprogram*}


The list of variables in a type expression can be computed by the following
Haskell function.

\begin{smallprogram*}
\> vars :: Type $\rightarrow$ [String] \\
\> vars ty = nub (v ty)  \\
\>     where v (TVar x) = [x] \\
\>           v (BinType op t1 t2) = v t1 ++ v t2 \\
\end{smallprogram*}

Note that {\it{nub}} is in the library {\em{Data.List}} and eliminates
duplicate entries in a list.


\section{Substitutions}

A {\em{type substitution }} is a function of type ${\cal{VT}}\rightarrow{}T$
mapping type variables to types.  Since the Haskell datatype {\it{Type}} uses
{\tt{String}} to represent variables substitutions can be defined in Haskell by
the following type:
\begin{smallprogram*}
\> Substitution :: String $\rightarrow$ Type \\
\end{smallprogram*}
We define the following function to recursively apply a substitution to a type.
\begin{smallprogram*}
\> subst :: Substitution $\rightarrow$ Type $\rightarrow$ Type\\
\> subst s (TVar x) = s x \\
\> subst s (BinType op t1 t2) = BinType op (subst s t1) (subst s t2) \\
\end{smallprogram*}
Note that applying a substitution to a type can only add structure at the
leaves {\em{i.e.}} applying a substitution can not change the top-level shape of
the syntax tree of the type, it can only change a variable which is a leaf.

The identity substitution is the one that maps strings {\it{x}} to types of the tome {\it{TVar x}}.
\begin{smallprogram*}
\> idSubst :: Substitution\\
\> idSubst x = (TVar x)\\
\end{smallprogram*}
We have the following theorem.
\begin{theorem}
\[\forall{}t:{\mathit{Type}}.\;\; {\mathit{subst}} \; {\mathit{idSubst}} \; t = t\]
\end{theorem}
You could prove this theorem by induction on the structure of the type $t$.

We can write down substitutions by enumerating the points where they differ
from {\it{idSubst}}.  For example the substitution that behaves like
{\it{idSubst}} except on variables $\alpha$ and $\beta$  could be written as
\[s = \{\alpha\mapsto{}\tau_1,\beta\mapsto\tau_2\}\]

We define the pointwise update of a function as follows:
\begin{smallprogram*}
\> update (x,v) f = ($\backslash\,$y $\rightarrow$  if y == x then v else f y)\\
\end{smallprogram*} 

Thus, the substitution $s$ enumerated above could be computed in Haskell by the following expression:
\begin{smallprogram*}
\> update ($\beta$, $\tau_2$) (update ($\alpha$,$\tau_2$) idSubst) \\
\end{smallprogram*} 



\section{Unification}

Given a pair of types $\tau_1$ and $\tau_2$ we say they are {\em{unifiable}} if
there is a substitution (call it $\sigma$) such that
\[{\mathit{subst\ }} \sigma \;\tau_1 = {\mathit{subst\ }} \sigma\; \tau_2\]

For example, the types $\alpha\rightarrow\alpha$ and
$\beta\times\gamma\rightarrow\delta$ is unifiable by a substitution of the
following form:
\[\begin{array}{l}
\{\alpha \mapsto  \beta\times\gamma,\; \delta \mapsto \alpha\}
\end{array}\]
Type expressions that do not share the same top level shape can not be unified.
For example, there is no substitution that unifies $\alpha\rightarrow\beta$
with $\alpha\times\beta$ because $\rightarrow$ and $\times$ do not match.
There is also a problem with a case like $\alpha$ and $\alpha\rightarrow\beta$.
If you try to figure out a way to do this, you see that $\alpha$ must become
something like $\alpha\rightarrow\beta$.  Consider what happens if we apply the
substitution $s = \{\alpha\mapsto(\alpha\rightarrow\beta)\}$.
\[\begin{array}{l}
\> {\mathit{subst}}\;s\;\alpha = (\alpha\rightarrow\beta)\\
\> {\mathit{subst}}\;s\; (\alpha\rightarrow\beta) = (\alpha\rightarrow\beta)\rightarrow\beta\\
\end{array}\]
So this leads to a kind of loop.  We keep expanding $\alpha$ to a term that has
$\alpha$ in it so we'll never get a match on both sides. This is called an
{\em{occurs check failure}}.


The unification algorithm takes two terms and, if they are unifiable, returns a
substitution that will make them identical.  Here is the unification algorithm
described mathematically where $\otimes$ is one of the type constructors
$\{\rightarrow,\times\}$.

\begin{smallprogram*}
\> unify $\alpha$ $\alpha$ = idSubst \\
\> unify $\alpha$ $\beta$ = update ($\alpha, \beta$) idSubst \\
\> unify $\alpha$ $(\tau_1\otimes\tau_2)$ = \\
\>             if $\alpha\in$ vars$(\tau_1\otimes\tau_2)$ then\\
\>                error "Occurs check failure"\\
\>             else\\
\>                update ($\alpha, \tau_1\otimes\tau_2$) idSubst \\
\> unify $(\tau_1\otimes\tau_2)$ $\alpha$ = unify $\alpha$ $(\tau_1\otimes\tau_2)$ \\
\> unify $(\tau_1\otimes_1\tau_2)$ $(\tau_3\otimes_2\tau_4)$ = if $\otimes_1$ == $\otimes_2$ then\\
\>                                       (subst s2) . s1 \\
\>                                     else \\
\>                                       error "not unifiable." \\
\>               where s1 = unify $\tau_1$ $\tau_3$  \\
\>                       s2 = unify (subst s1 $\tau_2$) (subst s1 $\tau_4$) \\
\end{smallprogram*}

\exercise{Using the base code provided on the web-page implement the
{\it{unify}} function in Haskell.}


\end{document}
% Local Variables:
% mode:latex
% comment-column:0
% comment-start: "% "
% compile-command: "pdflatex hw11c"
% fill-column:79
% End:




