\documentclass[11pt]{article}

\include{header}

\begin{document}
\homework{12}{5 October}
\exercise{ Read sections 4.2, 4.3, 4.4 and 4.5 of Bird (pp. 95---128)}

\exercise{Write a recursive function to implement {\tt{ordered1}} which tells
whether a list is in strictly increasing order or not.
\begin{program*}
\>    ordered1 :: (Ord a) => [a] -> Bool
\end{program*}
Write enough test cases to convince the grader that your code is correct.
}

\exercise{ There is a function {\tt{f}} such that the following definition
behaves as the predicate {\tt{ordered1}} defined above.  Fill in the ``???''.

\begin{program*}
\>    pairs [] = []\\
\>    pairs m = zip m (tail m) \\
\> \\
\>    ordered2 m = foldr f True (pairs m)  \\
\>      where f (x,y) z = ??? 
\end{program*}
To solve this problem you'll probably have to experiment with the function
{\tt{pairs}} to see what it does.

Write enough test cases to convince the grader that your code is correct.
}


\exercise{Implement the function {\tt{filter1}} which has the same type as, and
behaves just like {\tt{filter}} except that it is defined using {\tt{foldr}}.
Write enough test cases to convince the grader that your code is correct.
}


\exercise{Write Haskell code to implement {\tt{partition}} using the {\tt{foldr}} function.
\begin{program*}
\>    partition :: (a -> Bool) -> [a] -> ([a],[a])  
\end{program*}

The specification\footnote{This means that this equation characterizes
the {\tt{partition}} function, {\em{i.e.}} for any implementation of
partition, this property must hold.} for {\tt{partition}} is as
follows:
\begin{program*}
\>    partition p l = (filter p l, filter (not . p) l)
\end{program*}
Recall that {\tt{(f . g)}} denotes function composition, {\em{i.e.}} {\tt{(f . g)(x) = f(g(x))}}.

Write enough test cases to convince the grader that your code is correct.  You
can use the specification to test the output of your version for correctness.

}





\end{document}


% Local Variables:
% mode:latex
% comment-column:0
% comment-start: "% "
% compile-command: "pdflatex hw12b"
% fill-column:79
% End:




