\documentclass[11pt]{article}

\include{header}
\newcommand{\quot}[1]{"{\tt{{#1}}}"}
\newcommand{\mybs}{$\backslash$}

\begin{document}
\homework{14}{6 November}

The positive integers have representations as sequences of 1 or more
digits.  A negative integer has the form ``$(- k )$'' where $k$
is a positive integer.

\exercise{Write a parser {\tt{intp}} that parses integers of this form.
You can use the basic parsers provided in the file Parser.hs.
Here is some example behavior:

\begin{program*}
\> *Expr> :t intp \\
\> intp :: Parser Int \\
\> *Expr> apply intp "10" \\
\> [(10,"")] \\
\> *Expr> apply intp "01" \\
\> [(1,"")] \\
\> *Expr> apply intp "(-10)" \\
\> [(-10,"")] \\
\> *Expr> apply intp "(-10 " \\
\> [] \\
\> *Expr> apply intp " -10 " \\
\> [] \\
\> *Expr> apply intp "0000" \\
\> [(0,"")] \\
\end{program*}
}

\section{Expressions}

Consider the following concrete grammar for expressions.

\begin{program**}
\> expr    ::=  term ( '+' expr $\mid$ $\epsilon$) \\
\> term    ::=  factor ( '*' expr $\mid$ $\epsilon$) \\
\> factor    ::=  '(' expr ')' $\mid$ nat \\
\end{program**}

Here is a parser that calculates the values of expressions in this language.

\begin{program*}
\>  expr :: Parser Int  \\
\>  expr = do t <- term  \\
\>            do symbol "+"  \\
\>               e <- expr  \\
\>               return (t + e)  \\
\>              +++ return t  \\
\>    \\
\>  term :: Parser Int  \\
\>  term = do f <- factor  \\
\>            do symbol "*"  \\
\>               t <- term   \\
\>               return (f * t)  \\
\>              +++ return f  \\
\>    \\
\>  factor :: Parser Int  \\
\>  factor = do symbol "("  \\
\>              e <- expr  \\
\>              symbol ")"  \\
\>              return e  \\
\>             +++ natural  \\
\end{program*}

Now, consider the following expression datatype.

\begin{program*}
\> data BinOp = Add | Times \\
\>    deriving Show \\
\>  \\
\> data Exp = Const Int | BinExp BinOp Exp Exp  \\
\>    deriving Show  \\
\>  \\
\end{program*}


\exercise{Write parsers {\tt{expr', term', }} and {\tt{factor'}} of type
  {\tt{Parser Expr}} for parsing strings into the {\tt{Expr}} type. 

Here are some example runs:

\begin{program*}
\>  *Expression> parse expr' "55"  \\
\>  parse expr' "55"  \\
\>  [(Const 55,"")]  \\
\>  *Expression> parse expr' "55 + 23"  \\
\>  parse expr' "55 + 23"  \\
\>  [(BinExp Add (Const 55) (Const 23),"")]  \\
\>  *Expression> parse expr' "55 + 23 + 24"  \\
\>  parse expr' "55 + 23 + 24"  \\
\>  [(BinExp Add (Const 55) (BinExp Add (Const 23) (Const 24)),"")]  \\
\>  *Expression> parse expr' "55 + 23 + 24 * 25"  \\
\>  parse expr' "55 + 23 + 24 * 25"  \\
\>  [(BinExp Add (Const 55) (BinExp Add (Const 23) (BinExp Times (Const 24) (Const 25))),"")]  \\
\>  *Expression> parse expr' "55 + (23 + 24) * 25"  \\
\>  parse expr' "55 + (23 + 24) * 25"  \\
\>  [(BinExp Add (Const 55) (BinExp Times (BinExp Add (Const 23) (Const 24)) (Const 25)),"")]  \\
\>  *Expression>   \\
  \end{program*}  

\end{document}


% Local Variables:
% mode:latex
% comment-column:0
% comment-start: "% "
% compile-command: "pdflatex hw14d"
% fill-column:79
% End:




