\documentclass[11pt]{article}

\include{header}

\begin{document}
\homework{19}{20 November}


\exercise{ Read chapter 10 in Bird.}

\section{Type Derivations}


\subsection{Proof Rules}

Sequents in the system (which represent the state of a type derivation) are of
the form:

\[\Gamma,E \vdash M : T\]
In this structure, $\Gamma$ is a {\em{context}} representing a state of
knowledge about the types of some variables.  Contexts have the form:
 \[\Gamma = [x_1:\tau_1,\cdots{},x_k:\tau_k]\]
where the $x_i$'s are variables and $\tau_i$'s are types.

$E$ is a list of constraints between pairs
of types and in the rules is presented as follows:
\[E=\{\tau_{(1,1)} = \tau_{(1,2)}, \cdots , \tau_{(k,1)} = \tau_{(k,2)}\}\]
wher $\tau_{i,j}$'s are types.  

We write $\Gamma\backslash{}x$ to denote the list obtained from $\Gamma$ by
deleting all pairs whose first element is $x$.

As presented in the last homework, The proof rules for Wand's type
inference system including product types are given as follows:


\AxiomRule{\Gamma,\{\alpha=\tau\}}{x:\tau}{Ax} {\hspace{.25in}} if $(x,\alpha)\in\Gamma$.
\vspace{.125in}\\

\SequentRule{\Gamma,E\cup{}\{\tau=\alpha\rightarrow\beta\}}{\lambda{}x.M : \tau}{[x:\alpha]{\tt{++}}(\Gamma\backslash{x}),\,E}{M:\beta}{Abs}
{\hspace{.25in}} where $\alpha$ and $\beta$ are fresh.
\vspace{.125in}\\

\SequentRuleTwo{\Gamma,E_1\cup{}E_2}{MN:\tau}{\Gamma,E_1}{M:\alpha\rightarrow\tau}{\Gamma,E_2}{N:\alpha}{App}
{\hspace{.25in}} where $\alpha$ is fresh.
\vspace{.125in}\\

\SequentRuleTwo{\Gamma,E_1\cup{}E_2\cup{}\{\tau=\alpha\times\beta\}}{\pair{M,N} : \tau}{\Gamma,E_1}{M:\alpha}{\Gamma,E_2}{N:\beta}{Pair}
{\hspace{.25in}} where $\alpha$ and $\beta$ are fresh.
\vspace{.125in}\\

\SequentRuleTwo{\Gamma,E_1\cup{}E_2}{\spread{M}{x,y}{N}:\tau}{\Gamma,E_1}{M:\alpha\times\beta}{\{x:\alpha,y:\beta\}\cup((\Gamma\backslash{}x)\backslash{}y),E_2}{N:\tau}{Spread}
{\hspace{.25in}}
\begin{tabular}{l}
 where $\alpha$ and \\
 $\beta$ are fresh.
\end{tabular}
\vspace{.125in}\\

In the last homework, some sample deriviations were given.  Recall that to
prove a sequent of the form $\Gamma,E\vdash{}M:\tau$ you leave $E$ blank and
work backwards (up) until you get to tghe leaves and then compute the
constraint sets $E$ coming back down through the derivation tree.  I strongly
urge you to study the examples in HW 18.

\exercise{
Complete the following devivation trees and compute the constraint sets $E$.

\[\begin{array}{ll}
i.) & [z:(\alpha\rightarrow\alpha)\rightarrow\beta],E\vdash z(\lambda{}w.w):\tau \\
ii.) &  [y:\beta],E\vdash (\lambda{}x.x)y : \tau \\
iii.) &  [x:\alpha,y:\beta],E\vdash spread(\pair{x,y};z,w.\pair{w,z}): \tau \\
\end{array}
\]



}


\end{document}


% Local Variables:
% mode:latex
% comment-column:0
% comment-start: "% "
% compile-command: "pdflatex hw19a"
% fill-column:79
% End:




