\documentclass[11pt]{article}

\include{header}

\begin{document}
\homework{2}{30 August}

\section{}

\problem{Read chapter 1 of Bird -- since the book is not in yet, I have linked to a pdf copy of that chapter on the course web-page.}


\section{}
Consider the following code.

\begin{program*}

\> plus :: (Integer, Integer) -> Integer \\
\> plus (x,y) = x + y \\
\>   \\\
\> cplus :: Integer -> Integer -> Integer \\
\> cplus x y = x + y \\
\end{program*}


The function {\tt{plus}} takes its arguments all at once packaged in a pair
while {\tt{cplus}} takes its arguments one at a time.

In class, we discussed how Haskell supports a notation for describing a
function without forcing you to choose a name for it.

The general form is 
\[ \backslash x \;\;\rightarrow\:\; e \]
where $x$ is a variable and $e$ is a Haskell expression.

Note that in Haskell ``$\rightarrow$'' is used to denote the {\em{type
constructor}} for functions ({\em{e.g.}} if $\gamma$ and $\delta$ are types,
then $\gamma \rightarrow \delta$ is the type of functions from $\gamma$ to
$\delta$.  Also, ``$\rightarrow$'' is used in the expression language to
describe an actual function, $(\backslash x \rightarrow e)$ denotes a function
whose single argument is referred to in the expression $e$ by the variable $x$.

This overloading of syntax is similar to that for Cartesian products.  If
$\gamma$ and $\delta$ are types then $(\gamma,\delta)$ is the type whose
elements are the pairs where the first element comes from $\gamma$ and the
second element comes from $\delta$. But also, if $a\in\gamma$ and $b\in\delta$,
then the pair $(a,b)\in(\gamma,\delta)$.  So the developers of Haskell have
used the same notation for the type constructor and to construct the elements
of the type in both cases.


Now, consider the following interaction with the Haskell interpreter.
\begin{program*}
  \> Main> :t cplus    \\
  \> cplus :: Integer -> Integer -> Integer    \\
  \> Main> :t cplus 7    \\
  \> cplus 7 :: Integer -> Integer    \\
\end{program*}
Evidently {\tt{cplus 7}} is a function of type {\tt{Integer -> Integer}}.  But
what function is it?  It is the function that is expecting an input {\tt{y}}
and will compute the expression {\tt{ 7 + y}}.  So, it is the function
described by the following expression:\\ \begin{center}{\tt{$\backslash$y -> 7
+ y}} \end{center} This form of Haskell expression is called a
{\em{lambda-term}} ($\lambda$-term).

We can write $(\backslash x \; y \rightarrow e)$  for  $(\backslash x \rightarrow (\backslash y \rightarrow e))$.

Every function can be written in a form where no arguments are declared on the
left side of the definition.  If $e$ is an arbitrary Haskell expression, then
the following examples show how this works.
\[\begin{array}{lclcl}
f\, x = e & {\rm{is\ the\ same\ as }} & f = \backslash{}x \rightarrow e \\
g \, x \, y = e & {\rm{is\ the\ same\ as }} & g \, x = \backslash{}y \rightarrow e & {\rm{is\ the\ same\ as }} & g = \backslash{}x\, y \rightarrow e \\
\end{array}\]


We say that {\tt{cplus}} is in {\em{Curried Form}}.  In this form of function
definition, where the function takes its arguments one at a time, is named
after Haskell Curry (1900-1982), an American mathematician and logician.  As
you might guess, the Haskell programming language is named after him as well.
A Curried function  takes its arguments one at a time -- as opposed to packaged up in a pair.


Consider the following function\footnote{Note that curry and uncurry are
defined in the standard prelude so if you want to try your own version give it
a name like {\tt{curry1}} or {\tt{curry'}}.} called {\tt{curry}}:
\begin{program*}
  \> curry f  =  $\backslash$x -> $\backslash$y -> f(x,y)
\end{program*}


We can also move the arguments over to the left side of the ``='' and use the
entirely equivalence definition
\begin{program*}
  \> curry f x y  = f(x,y)
\end{program*}


In Haskell we have the following.

\begin{program*}
  \> Main> :t curry \\
  \> curry :: ((a,b) -> c) -> a -> b -> c \\
  \> Main>  \\
\end{program*}
So {\tt{curry}} takes a function of type {\tt{((a,b) -> c)}} and returns a function of type {\tt{(a -> b -> c)}}.

Also, consider the following definition: %

\begin{program*}
\> uncurry f = $\backslash$(x,y) ->  f x y    \\
\end{program*}

An equivalent definition that does not contain any lambda terms is given as
follows:

\begin{program*}
\> uncurry f (x,y) = f x y    \\
\end{program*}


Here is the type of {\tt{uncurry}}.

\begin{program*}
\> Main> :t uncurry \\ 
\> uncurry :: (a -> b -> c) -> (a,b) -> c \\ 
\> Main>  \\ 
\end{program*}


The function {\tt{curry}} works on functions expecting two arguments packaged
 up as a pair and returns a function that expects arguments one at a time. The
 {\tt{uncurry}} functions works on functions expecting two arguments.  What are
 the analogues of curry and uncurry for functions of three arguments?

\problem{Write the function {\tt{curry3}} having the following  type:

\begin{program*}
\> Main> :t curry3 \\ 
\> :t curry3 \\ 
\> curry3 :: ((a,b,c) -> d) -> a -> b -> c -> d \\ 
\> Main>  \\ 
\end{program*}
}

\problem{Write the function {\tt{uncurry3}} having the following  type:

\begin{program*}
\> Main> :t uncurry3 \\ 
\> :t uncurry3 \\ 
\> uncurry3 :: (a -> b -> c -> d) -> (a,b,c) -> d \\
\> Main>  \\ 
\end{program*}

\problem{Test you code to get the results in the hw2\_expected.txt output (linked onthe webpage.)}

\end{document}
% Local Variables:
% mode:latex
% comment-column:0
% comment-start: "\\> "
% comment-end: "\\\\ "
% compile-command: "pdflatex hw2c"
% fill-column:79
% End:




