\documentclass[11pt]{article}

\include{header}

\begin{document}
\homework{4}{12 September}

\section{}

\problem{When you get your books, read chapters 1 and 2.}

If you have not ordered a copy of the book yet -- you
should.  

\section{}
In class we discussed the following code.

\begin{program*}

\> plus :: (Integer, Integer) -> Integer \\
\> plus (x,y) = x + y \\
\>   \\\
\> plusc :: Integer -> Integer -> Integer \\
\> plusc x y = x + y \\
\end{program*}


The function {\tt{plus}} takes its arguments all at once packaged in a pair
while {\tt{plusc}} takes its arguments one at a time.

We discussed how Haskell supports a notation for describing a function without
forcing you to choose a name for it.

The general form is 
\[ \backslash x \;\;\rightarrow\:\; e \]
where $x$ is a variable and $e$ is a Haskell expression.

Note that in Haskell ``$\rightarrow$'' is used to denote the {\em{type
constructor}} for functions ({\em{e.g.}} if $\gamma$ and $\delta$ are types,
then $\gamma \rightarrow \delta$ is the type of functions from $\gamma$ to
$\delta$.  Also, ``$\rightarrow$'' is used in the expression language to
describe an actual function, $(\backslash x \rightarrow e)$ denotes a function
whose single argument is referred to in the expression $e$ by the variable $x$.

This overloading of syntax is similar to that for Cartesian products.  If
$\gamma$ and $\delta$ are types then $(\gamma,\delta)$ is the type whose
elements are the pairs where the first element comes from $\gamma$ and the
second element comes from $\delta$. But also, if $a\in\gamma$ and $b\in\delta$,
then the pair $(a,b)\in(\gamma,\delta)$.  So the developers of Haskell have
used the same notation for the type constructor and to construct the elements
of the type in both cases.


Now, consider the following interaction with the Haskell interpreter.
\begin{program*}
  \> Main> :type plusc    \\
  \> :type plusc    \\
  \> plusc :: Integer -> Integer -> Integer    \\
  \> Main> :type plusc 7    \\
  \> :type plusc 7    \\
  \> plusc 7 :: Integer -> Integer    \\
\end{program*}
Evidently {\tt{plusc 7}} is a function of type {\tt{Integer -> Integer}}.  But
what function is it?  It is the function that is expecting an input {\tt{y}}
and will compute the expression {\tt{ 7 + y}}.  So, it is the function
described by the following expression:\\ \begin{center}{\tt{$\backslash$y -> 7
+ y}} \end{center} This form of Haskell expression is called a
{\em{lambda-term}} ($\lambda$-term).

We can write $(\backslash x \; y \rightarrow e)$  for  $(\backslash x \rightarrow (\backslash y \rightarrow e))$.

Every function can be written in a form where no arguments are declared on the
left side of the definition.  If $e$ is an arbitrary Haskell expression, then
the following examples show how this works.
\[\begin{array}{lclcl}
f\, x = e & {\rm{is\ the\ same\ as }} & f = \backslash{}x \rightarrow e \\
g \, x \, y = e & {\rm{is\ the\ same\ as }} & g \, x = \backslash{}y \rightarrow e & {\rm{is\ the\ same\ as }} & g = \backslash{}x\, y \rightarrow e \\
\end{array}\]


We say that {\tt{plusc}} is in {\em{Curried Form}}\footnote{``Currying'' is
the verb form of the adjective ``Curried''.  This form of function definition,
where the function takes its arguments one at a time, is named after Haskell
Curry (1900-1982), an American mathematician and logician.  As you might guess,
the Haskell programming language is named after him as well.}; that is, it
takes its arguments one at a time.


Consider the following function:
\begin{program*}
  \> curry f x y  =  f(x,y)
\end{program*}

In Haskell we have the following.

\begin{program*}
  \> Main> :l test \\
  \> :l test \\
  \> Main> :t curry \\
  \> :t curry \\
  \> curry :: ((a,b) -> c) -> a -> b -> c \\
  \> Main>  \\
\end{program*}
So {\tt{curry}} takes a function of type {\tt{((a,b) -> c)}} and returns a function of type {\tt{(a -> b -> c)}}.

So, consider the following definitions:

\begin{program*}
\>   \\
\> curry f x y = f (x,y)    \\
\>     \\
\> uncurry f (x,y) =  f x y    \\
\>     \\
\> flip f x y = f y x    \\
\>     \\
\> flop f (x,y) = f (y,x)    \\
\>     \\
\> minus (x,y) = x - y    \\
\>     \\
\end{program*}


Using calculational methods we can prove, for example the following:

\[ flop \; minus\; (x,y) =  (uncurry\;  . \; flip \; . \; curry)\; plus\; x \; y \]
{\bf{proof:} }

{\tt{
\begin{tabular}{ll}
     flop minus (x,y) {\mbox{\hspace{1in}}}   &  \hfill (uncurry . (flip . curry)) minus (x,  y)\\
    \pnote{by  definition of flop}            &  \pnote{by  definition of compose} \\
     =  minus(y,x)                            &  =  uncurry  (flip . curry) minus (x, y)) \\
                                              & \pnote{by associativity of application} \\
                                              &  =  uncurry ((flip . curry) minus) x y \\
                                              & \pnote{by definition of compose} \\
                                              &  =  uncurry ((flip . curry) minus) x y \\

                                              & \pnote{by definition of flip}\\
                                              &  =  uncurry  (curry minus) y x) \\
\end{tabular}
}}




\problem{
You may prove the following, or (if you don't want to do the proofs), run extensive tests to show the following hold.

\[\begin{array}{ll}
i.) & uncurry \; plusc = plus \\
ii.) & curry \; plus = plusc \\
iii.) & curry (uncurry \; plusc) = plusc \\
iv.) & uncurry (curry \; plus) = plus \\

\end{array}\]
}



\problem{The built in function {\tt{flip :: (a -> b -> c) -> b -> a -> c}}
swaps the order of the arguments to a two argument curried function. Thus,
\[flip f x y = f y x \]

Write a function {\tt{flop :: ((a,b) -> c) -> (b,a) -> c}} and show that 

  }



\end{document}
% Local Variables:
% mode:latex
% comment-column:0
% comment-start: "% "
% compile-command: "latex hw3a; dvips hw3a -o hw3a.ps; pdflatex hw3a"
% fill-column:79
% End:




