\documentclass[11pt]{article}

\include{header}

\begin{document}
\homework{3}{4 September}

\section{}

\problem{When you get your books, read chapters 1 and 2.}

If you have not ordered a copy of the book yet -- you
should.  

\section{}
In class we discussed the following code.

\begin{program*}

\> plus :: (Integer, Integer) -> Integer \\
\> plus (x,y) = x + y \\
\>   \\\
\> plusc :: Integer -> Integer -> Integer \\
\> plusc x y = x + y \\
\end{program*}


The function {\tt{plus}} takes its arguments all at once packaged in a pair
while {\tt{plusc}} takes its arguments one at a time.

We discussed how Haskell supports a notation for describing a function without
forcing you to choose a name for it.

The general form is 
\[ \backslash x \;\;\rightarrow\:\; e \]
where $x$ is a variable and $e$ is a Haskell expression.

Note that in Haskell ``$\rightarrow$'' is used to denote the {\em{type
constructor}} for functions ({\em{e.g.}} if $\gamma$ and $\delta$ are types,
then $\gamma \rightarrow \delta$ is the type of functions from $\gamma$ to
$\delta$.  Also, ``$\rightarrow$'' is used in the expression language to
describe an actual function, $(\backslash x \rightarrow e)$ denotes a function
whose single argument is referred to in the expression $e$ by the variable $x$.

This overloading of syntax is similar to that for Cartesian products.  If
$\gamma$ and $\delta$ are types then $(\gamma,\delta)$ is the type whose
elements are the pairs where the first element comes from $\gamma$ and the
second element comes from $\delta$. But also, if $a\in\gamma$ and $b\in\delta$,
then the pair $(a,b)\in(\gamma,\delta)$.  So the developers of Haskell have
used the same notation for the type constructor and to construct the elements
of the type in both cases.


Now, consider the following interaction with the Haskell interpreter.
\begin{program*}
  \> Main> :type plusc    \\
  \> plusc :: Integer -> Integer -> Integer    \\
  \> Main> :type plusc 7    \\
  \> plusc 7 :: Integer -> Integer    \\
\end{program*}
Evidently {\tt{plusc 7}} is a function of type {\tt{Integer -> Integer}}.  But
what function is it?  It is the function that is expecting an input {\tt{y}}
and will compute the expression {\tt{ 7 + y}}.  So, it is the function
described by the following expression:\\ \begin{center}{\tt{$\backslash$y -> 7
+ y}} \end{center} This form of Haskell expression is called a
{\em{lambda-term}} ($\lambda$-term).

We can write $(\backslash x \; y \rightarrow e)$  for  $(\backslash x \rightarrow (\backslash y \rightarrow e))$.

Every function can be written in a form where no arguments are declared on the
left side of the definition.  If $e$ is an arbitrary Haskell expression, then
the following examples show how this works.
\[\begin{array}{lclcl}
f\, x = e & {\rm{is\ the\ same\ as }} & f = \backslash{}x \rightarrow e \\
g \, x \, y = e & {\rm{is\ the\ same\ as }} & g \, x = \backslash{}y \rightarrow e & {\rm{is\ the\ same\ as }} & g = \backslash{}x\, y \rightarrow e \\
\end{array}\]


We say that {\tt{plusc}} is in {\em{Curry Form}}.  In this form of function
definition, where the function takes its arguments one at a time, is named
after Haskell Curry (1900-1982), an American mathematician and logician.  As
you might guess, the Haskell programming language is named after him as well.
It takes its arguments one at a time -- as opposed to packaged up in a pair.


Consider the following function (as defined in class):
\begin{program*}
  \> curry f x y  =  f(x,y)
\end{program*}

In Haskell we have the following.

\begin{program*}
  \> Main> :t curry \\
  \> curry :: ((a,b) -> c) -> a -> b -> c \\
  \> Main>  \\
\end{program*}
So {\tt{curry}} takes a function of type {\tt{((a,b) -> c)}} and returns a function of type {\tt{(a -> b -> c)}}.

Also, consider the following definition: %

\begin{program*}
\> uncurry f p =  f (fst p) (snd p)    \\
\end{program*}

Or equivalently, 

\begin{program*}
\> uncurry f (x,y) =  f x y    \\
\end{program*}

Here is the type of {\tt{uncurry}}.

\begin{program*}
\> Main> :t uncurry \\ 
\> uncurry :: (a -> b -> c) -> (a,b) -> c \\ 
\> Main>  \\ 
\end{program*}


Now, these functions work on functions of 2 arguments.  What about functions of three arguments?

\problem{Write the function {\tt{curry3}} having the following  type:

\begin{program*}
\> Main> :t curry3 \\ 
\> :t curry3 \\ 
\> curry3 :: ((a,b,c) -> d) -> a -> b -> c -> d \\ 
\> Main>  \\ 
\end{program*}
}

\problem{Write the function {\tt{uncurry3}} having the following  type:

\begin{program*}
\> Main> :t uncurry3 \\ 
\> :t uncurry3 \\ 
\> uncurry3 :: (a -> b -> c -> d) -> (a,b,c) -> d \\
\> Main>  \\ 
\end{program*}

HINT: You might want to use the second definition of uncurry as a model (where the pair is destructured in the argument).  As an experiment you might try to evaluate the term {\tt{snd (1,''xxx'',[])}}.

}
\end{document}
% Local Variables:
% mode:latex
% comment-column:0
% comment-start: "\\> "
% comment-end: "\\\\ "
% compile-command: "pdflatex hw3d"
% fill-column:79
% End:




