\documentclass[11pt]{article}
\usepackage{fullpage}
\include{header}

\begin{document}
\homework{4}{7 September}

\section{}

\problem{Read Chapter 1 and 2 of the Bird text.}

\section{info, infix operators and precedence}

GHC will tell you the type, where the function is defined and, for infix
operators, the precedence of the operator.  Here are some examples.

\begin{program*}
\> *Main> :info (\&\&) \\
\> (\&\&) :: Bool -> Bool -> Bool    -- Defined in GHC.Base \\
\> infixr 3 \&\& \\
\> *Main> :info (||) \\
\> (||) :: Bool -> Bool -> Bool    -- Defined in GHC.Base \\
\> infixr 2 || \\
\> *Main> :info (.) \\
\> (.) :: (b -> c) -> (a -> b) -> a -> c   -- Defined in GHC.Base \\
\> infixr 9 . \\
\> *Main> :l infinity \\
\> :l infinity \\
\> [1 of 1] Compiling Main             ( infinity.hs, interpreted ) \\
\> Ok, modules loaded: Main. \\
\> *Main> :info infinity \\
\> :info infinity \\
\> infinity :: t   -- Defined at infinity.hs:1:0-7 \\
\> *Main>  \\
\end{program*}

Precedence level is an integer from 0 to 9 with (9 being the highest
precedence.)  Function application associates to the left and has precedence 10
so, for example {\tt{f x y + 7}} is {\tt{((f x) y) + 7}}.

Haskell allows users to declare your own infix operators and whether they
associate to the left ({\tt{infixl}}) or right ({\tt{infixr}}) or do not
associate ({\tt{infix}})).  You must also specify the precedence of the new
operator.

For example, in class I presented the definitions of conjunction ({\tt{\&\&}})
and disjunction ({\tt{||}}) as follows:

\begin{program*}
\> False \&\&  x = False \\
\> True \&\&  x = x \\
\> \\
\> False || x = x \\
\> True || x = True \\
\end{program*}
This code will load, but when you try to use it you will need to specify which
definition of {\tt{\&\&}} you mean to use.  \newpage{}For example, after loading the code
above I get the following behavior.
\begin{program*}
\> *Main> True \&\& False \\
\> True \&\& False \\
\>  \\
\> <interactive>:1:5: \\
\>     Ambiguous occurrence `\&\&' \\
\>     It could refer to either `Main.\&\&', defined at hw4c.hs:7:6 \\
\>                           or `Prelude.\&\&', imported from Prelude
\end{program*}

The error message is indicating that the name {\tt{\&\&}} is ambiguous. You
could fix this by specifying which of the two definitions of {\tt{\&\&}} you
intend.
\begin{program*}
\> *Main> True Main.\&\& False \\
\> False
\end{program*}

To avoid conflict with the built-in functions we might have written the
following code which declares two new infix operators that associated to the
right.

\begin{program*}
\> infixr 3 \&\&\& \\
\> (\&\&\&) :: Bool -> Bool -> Bool \\
\> False \&\&\& y = False \\
\> True \&\&\& y = y \\
\>  \\
\> infixr 2 ||| \\
\> (|||) :: Bool -> Bool -> Bool \\
\> False ||| y = y \\
\> True ||| y = True
\end{program*}

Now we can compute as we expect.
\begin{program*}
\> *Main> True \&\&\& False \\
\> False
\end{program*}



\begin{problem}
Rewrite\footnote{This is essentially problem 2.2.1 from Bird on page 34.} the definitions of {\tt{\&\&\&}} and {\tt{|||}} so that they behave just
like {\tt{\&\&}} and {\tt{||}} but only using {\tt{if-then-else}} and the
Boolean constants. Your definitions should look something like the following:
\begin{program*}
\> x \&\&\& y = ???  \\
\> x ||| y = ???
\end{program*}
The right side of your definition ({\tt{???}}) must only use the constructs
{\tt{True}}, {\tt{False}}, {\tt{if-then-else}}, and the variables {\tt{x}} and
{\tt{y}}.
\end{problem}

\begin{problem}
Declare two new infix operators implementing logical implication ({\tt{==>}})
and exclusive-or ({\tt{<+>}}). (If you do not know the truth table for
implication and exclusive-or, look at the expected output file.) Both operators
should associate to the right.  Implication should have precedence 0 (so that
it binds weaker than the other connectives) and exclusive-or should have the
same precedence as regular disjunction ({\tt{||}}).
\end{problem}


Check the linked expected output file to see the tests you should run.



\end{document}
% Local Variables:
% mode:latex
% comment-column:0
% comment-start: "% "
% compile-command: "pdflatex hw4c"
% fill-column:79
% End:




