\documentclass[11pt]{article}

\include{header}

\begin{document}
\homework{4}{10 September}

\section{Equality of functions}
Recall from class that functions are equal if and only if they are equal on all
inputs (this equality is called extensionality.)

\begin{definition}{{\bf{(extensionality)}}}
If $f,g::a\rightarrow{}b$ then they are defined to be (extensionally) equal as follows:
\[ f = g \; \definedAs\; \forall{}x\!:\!a.\; f(x) = g(x) \]
\end{definition}

So, we can prove two functions $f$ and $g$ are equal by choosing an arbitrary
$x$ of type $a$ and showing $f(x) = g(x)$.

\ \\ For example, if $f(x) = |x|$ (the absolute value) and $g(x) = x$ then,
$f\not={}g$ when we consider them as functions in the type $\Int \rightarrow
\Int$ since $f(-2) = 2$ and $g(-2)=-2$.  But, if we think of these functions as
elements of $\nat\rightarrow\nat$, they are equal.  To see this, choose an
arbitrary $x\in\nat$ and argue that $f(x) = g(x)$ {\em{i.e.}}  that $|x|=x$.
But this is trivially true when $x \ge 0$, which follows because $x\in\nat$.  \
\\

Recall the following Haskell definitions.
\begin{definition}{plus}
\[\begin{array}{l}
plus :: (Integer,Integer) \rightarrow Integer \\
plus(x,y) = x + y 
\end{array}\]
\end{definition}


\begin{definition}{plusc}
\[\begin{array}{l}
plusc :: Integer \rightarrow (Integer \rightarrow Integer) \\
plusc \; x \; y = x + y \\
\end{array}\]
\end{definition}

\begin{definition}{curry}
\[\begin{array}{l}
curry :: ((a,b) \rightarrow c) \rightarrow (a \rightarrow (b \rightarrow c))\\
curry\; f\; x\; y = f\; (x,y) 
\end{array}\]
\end{definition}

\begin{definition}{uncurry}
\[\begin{array}{l}
uncurry :: (a \rightarrow (b \rightarrow c)) \rightarrow ((a,b) \rightarrow c) \\
uncurry \;f\; (x,y) = f\; x\; y 
\end{array}\]
\end{definition}

In class we proved the following theorem:

\begin{theorem}{}
\[curry\;\; plus = plusc\]
\Proof Note that both $curry\;plus$ and $plusc$ have the type
$Integer\rightarrow (Integer \rightarrow Integer)$ {\em{i.e.}} they are
functions mapping an $Integers$ to a function of type
$Integer\rightarrow{}Integer$.  This means we can use extensionality to prove
they are equal as functions.  We must show the following.
\[\forall{}x:Integer.\; curry\;\; plus\; x = plusc\; x\]
Assume $x$ is an arbitrary $Integer$. Then we must show
\[curry\;\; plus\; x = plusc\; x\]
But $curry\;\; plus \; x$ and $plusc\; x$ are functions of type
$Integer\rightarrow{}Integer$. To show they are equal we use extensionality a
second time, to show:
\[\forall{}y:Integer.\; curry\;\; plus\; x\; y = plusc\; x\; y\]
We choose an arbitrary $y$ of type $Integer$ and  show the following
\[curry\;\; plus\; x\; y = plusc\; x\; y\]
Starting with the left side of the equality we get the following:
\[curry\;\; plus\; x\; y \stackrel{\pair{\pair{def.\,of\, curry}}}{=} plus\; (x,y)   \stackrel{\pair{\pair{def.\,of\, plus}}}{=} x + y \]
On the right side of the equality, we have the following:
\[plusc\; x\; y \stackrel{def.\,of\, plusc}{=} x + y \]
Since both sides of the equality are equal to $x+y$ we see that the functions
are equal.  \qed
\end{theorem}
\ \\


\begin{problem}
Prove the following theorem using extensionality.

\begin{theorem}{\bf{[uncurry-plusc]}} 
\[uncurry\;plusc = plus \]
\end{theorem}

Hint: The functions $(uncurry\;plusc)$ and $plus$ have the type
$(Integer,Integer)\rightarrow{}Integer$.  Extensionality for functions $f$ and
$g$ of this type can most conveniently  be written as
\[\forall{}(x,y):(Integer,Integer).\; f (x,y) = g (x,y)\]

\end{problem}
\newpage
\section{Function Composition}


\goodbreak\noindent{}Now, consider the following two definitions.
\vspace{.06125in}

\begin{definition}{Function Composition}
\[\begin{array}{l}
compose :: (b\rightarrow{}c)\rightarrow (a\rightarrow{}b) \rightarrow (a \rightarrow c)\\
compose\; f\; g \; x = f (g\; x)
\end{array}
\]
\end{definition}

In Haskell, $(compose\; f\; g)$ is written $(f\;.\;g)$, we will write
$(f\circ{}g)$ here.

\begin{definition}{Identity function}
\[\begin{array}{l}
id :: a \rightarrow {}a\\
id\; x = x
\end{array}
\]
\end{definition}

\begin{theorem}{Compose-id-right}
\[\forall{}f:a\rightarrow{}b.\; (f\circ{} id) = f\]
\Proof
Choose an arbitrary function $f::a\rightarrow{}b$ and show that 
\[ f\circ{}id = f\]
Since $f$ has type $a\rightarrow{}b$ and $id$ has type $a\rightarrow{}a$ we can
see that $f\circ{}id$ has the same type. (why?)  We use extensionality to show
that these two functions are equal, {\em{i.e.}} we must show:
\[\forall{}x:a.\, (f\circ{}id)\,x =f\; x\]
Chose an arbitrary $x$ in type $a$ and show
\[(f\circ{}id)\,x =f\; x\]
By definition of compose and definition of $id$ we get the following sequence of equalities.
\[(f\circ{}id)x = compose\;f\;id\;x = f\,(id\;x) = f \; x\]
This completes the proof.
\qed

\end{theorem}


\begin{problem}
Prove the following theorem. 
\begin{theorem}{\bf{[compose-id-left]}} 
\[\forall{}f:a \rightarrow b.\;id\circ{} f = f\]
\end{theorem}

Hint: First argue that $id\circ{}f$ and $f$ have the same type (note that
$id::b\rightarrow{}b$) and then use extensionality.


\end{problem}





\end{document}
% Local Variables:
% mode:latex
% comment-column:0
% comment-start: "% "
% compile-command: "pdflatex hw4f"
% fill-column:79
% End:

