\documentclass[11pt]{article}

\include{header}

\begin{document}
\homework{6}{19 September}

\section{}

\problem{Read Chapter 2 of Bird.}

\ \\
Here is some Haskell code for a type {\tt{Boolean}} that does not interfere with the builtin type {\tt{Bool}}.  I renamed {\tt{not}} to {\tt{neg}} so it does not conflict with the {\tt{not}} in 
the Haskell type {\tt{Bool}}.


\begin{program*}
\>    \\
\> module Boolean where     \\
\>    \\
\> data Boolean = FF | TT             deriving (Eq,Ord,Enum,Show)   \\
\>    \\
\> neg :: Boolean -> Boolean   \\
\> neg FF  = TT   \\
\> not TT = FF   \\
\>    \\
\> (/$\bf\backslash$), ($\bf\backslash$/) :: Boolean -> Boolean -> Boolean   \\
\>    \\
\> FF /$\bf\backslash$ x = FF   \\
\> TT /$\bf\backslash$ x = x   \\
\>    \\
\> FF $\bf\backslash$/ x = x   \\
\> TT $\bf\backslash$/ x = TT   \\
\>    \\
\end{program*} 


\problem{Do problems 2.1.1, 2.1.3, 2.1.6 on pages 34 -- 35. For 2.1.3 use the operator {\tt{(.=>)}} instead of {\tt{(=>)}} or just use the prefix {\tt{implies}}.}

\section{Type Classes}
The deriving clause in the declaration of the type {\tt{Boolean}} generates
many usefull functions for the type. Here are the declarations of these classes.

\subsection{{\tt{Eq}} Typeclass}

Described on page 31 of Bird.

\begin{program*}
\> class  Eq a  where   \\
\>         (==), (/=)  ::  a -> a -> Bool   \\
\>    \\
\>         x /= y  = not (x == y)   \\
\>         x == y  = not (x /= y)   \\
\>    \\
\end{program*}



\subsection{{\tt{Ord}} Typeclass}

Described on pg 32 of Bird.

\begin{program*}
\> class  (Eq a) => Ord a  where   \\
\>     compare              :: a -> a -> Ordering   \\
\>     (<), (<=), (>=), (>) :: a -> a -> Bool   \\
\>     max, min             :: a -> a -> a   \\
\>    \\
\>     compare x y | x == y    = EQ   \\
\>                 | x <= y    = LT   \\
\>                 | otherwise = GT   \\
\>    \\
\>     x <= y  = compare x y /= GT   \\
\>     x <  y  = compare x y == LT   \\
\>     x >= y  = compare x y /= LT   \\
\>     x >  y  = compare x y == GT   \\
\>    \\
\>     -- Note that (min x y, max x y) = (x,y) or (y,x)   \\
\>     max x y | x <= y    =  y   \\
\>             | otherwise =  x   \\
\>     min x y | x <= y    =  x   \\
\>             | otherwise =  y   \\
\>    \\
\end{program*}

\subsection{{\tt{Enum}} Typeclass}

Described on page 38 -- 40 of Bird.

\begin{program*}
\> class  Enum a  where   \\
\>     succ, pred     :: a -> a   \\
\>     toEnum         :: Int -> a   \\
\>     fromEnum       :: a -> Int   \\
\>     enumFrom       :: a -> [a]            -- [n..]   \\
\>     enumFromThen   :: a -> a -> [a]       -- [n,n'..]   \\
\>     enumFromTo     :: a -> a -> [a]       -- [n..m]   \\
\>     enumFromThenTo :: a -> a -> a -> [a]  -- [n,n'..m]   \\
\>    \\
\>     -- Default declarations given in Prelude   \\
\>    \\
\>    \\
\end{program*}

\subsection{{\tt{Show}} Typeclass}

Described on page 52 -- 54 of Bird.

\begin{program*}
\> class  Show a  where   \\
\>     showsPrec :: Int -> a -> ShowS   \\
\>     show      :: a -> String    \\
\>     showList  :: [a] -> ShowS   \\
\>    \\
\>     showsPrec \underline{\hspace{1em}} x s   = show x ++ s   \\
\>     show x            = showsPrec 0 x ""   \\
\>     -- ... default decl for showList given in Prelude   \\
\end{program*}

To find out more, read section 6.3 of of {\tt{http://www.haskell.org/onlinereport/basic.html}}. 



\problem{Write expressions that exercise the functions {\tt{( $==$, $/\!=$,
$<$, $<=$, $>=$, $>$, succ, pred, fromEnum, toEnum, show, showList)}}  generated
automatically for the type {\tt{Boolean}} by the Haskell deriving clause {\tt{(Eq,Ord,Enum,Show)}}.}


\end{document}
% Local Variables:
% mode:latex
% comment-column:0
% comment-start: "% "
% compile-command: "latex hw6; dvips hw6 -o hw6.ps; pdflatex hw6"
% fill-column:79
% End:




