\documentclass[11pt]{article}

\include{header}

\begin{document}
\homework{6}{16 September}

\section{}

\problem{Start reading chapter 3 of Bird.}

Here is some code having to do with dates.

\begin{program*}
\> \\
\> -- Following British convention ... the numbers in a date represent (day,month,year)  \\ 
\> data Date = DMY (Int, Int, Int) \\ 
\> \\
\> leap y = (y `mod` 4 == 0) \&\& not(y `mod` 100 == 0 ) \\
     ||  (y `mod` 4 == 0) \&\& (y `mod` 100 == 0 )   \&\& (y `mod` 400 == 0) \\ 
\> \\
\> days\_in\_month y m  \\ 
\>    | has\_31                     = [1..31] \\ 
\>    | not (has\_31) \&\& m /= 2     = [1..30] \\ 
\>    | m == 2 \&\& not (leap y)     = [1..28] \\ 
\>    | m == 2 \&\& leap y           = [1..29] \\ 
\>  where  \\ 
\>      has\_31 = m `elem` [1,3,5,7,8,10,12] \\ 
\> \\ 
\> goodDate (DMY(d,m,y)) = d `elem` (days\_in\_month y m) \\ 
\> \\
\> dates\_in\_year y = 
\>   do m <- [1..12] \\ 
\>      d <- days\_in\_month y m \\ 
\>      return (DMY (d,m,y)) \\ 
\> \\
\> data Months = January | Feburary | March | April | May | June |July \\
               | August | September | October | November | December  \\ 
\>   deriving (Enum, Show) 
\end{program*}

\problem{This is essentially the harder version of problem 2.7. of Bird
pp. 55. Make the type {\tt{Data}} an instance of the type class {\tt{Show}} and
define a show function that prints dates in the following (ordinal) form. If a
date is invalid -- you code should raise an exception by calling {\tt{error
"bad date!"}}.  Test your program by evaluating the expression
{\tt{(dates\_in\_year 1956)}} and submitting the output with your code. Note
that my show function appends {\tt{"$\backslash$n"}} to the end of the date
string so that they print on their own lines.
\begin{program*}
\> Main> DMY (2,1,2008) \\ 
\> 2nd January 2008 \\ 
\> Main> DMY (29,2,2008) \\ 
\> 29th Feburary 2008 \\ 
\> Main> DMY (29,2,2007) \\ 
\> Program error: bad date! \\ 
\end{program*}
}


\end{document}
% Local Variables:
% mode:latex
% comment-column:0
% comment-start: "\\> "
% comment-end: "\\\\ "
% compile-command: "pdflatex hw6a"
% fill-column:79
% End:




