\documentclass[11pt]{article}

\include{header}

\begin{document}
\homework{6}{9 September}

\problem{If you haven't done so - read Chapter 1 (Starting Out) from LYAHFGG
  and read chapter 2 (Believe the Type).}  
\ \\

Note that the Haskell function {\em{error}} takes a string and halts the
computation using the string as the error message.

{\em{
\begin{tabular}{l}
Prelude$>$ error "Oops!" \\
*** Exception: Oops! \\
Prelude$>$ error "Splat!" \\
*** Exception: Splat!
\end{tabular}
}}
\problem{Write functions having the following types:  You will need to use recursion - and don't just use the built-in version if there is one.

\[\begin{array}{lcl}
last &::& [a] \rightarrow a\\
select & :: & [a] \rightarrow {} Int \rightarrow a\\
middle & :: & [a] \rightarrow a \\
split & :: & [a] \rightarrow Int \rightarrow ([a],[a])\\
repeat & :: & (a \rightarrow a) \rightarrow Int \rightarrow a \rightarrow a
\end{array}\]

The {\em{last}} function takes a list and returns the last element of the list
or calls {\em{error}} if the list is empty.  The function {\em{select}} takes a
list (say ${\mathit{xs}}$) and an integer (say $k$) and returns the $k^{th}$
element of the list ${\mathit{xs}}$ (using zero based indexing).  If $k<0$ or
$k \ge {\mathit{length}}\; {\mathit{xs}}$ then call {\em{error}}.  The
{\em{middle}} function takes a list and returns the middle element -- if the
list is of even length, you can implement your function to have a leftist or
rightist bias -- your choice. {em{split}} take s a list and a position $k$ in
the list and returns a pair of lists. The first element of the pair contains
the first $k$ elements of the input list and the second element of the pair
contains the the $k+1^{st}$ through the last element of the list.  It should be
that if you append the two output lists - you get back a list equal to the
input list.  A call {\em{repeat f k x}} applies the function $f$ $k$ times to
the input $x$. Thus \[repeat\; f\; k\; x = \stackrel{k\;times}{\overbrace{f (f
    \cdots (f} x))}\] If $k=0$ then the result is the identity function
{\em{id}}.

You should implement some tests to convince the grader your code works.


}

\end{document}

% Local Variables:
% mode:latex
% comment-column:0
% comment-start: "% "
% compile-command: "pdflatex hw6d"
% fill-column:79
% End:




