\documentclass[11pt]{article}

\include{header}
\newcommand{\sS}{{{Succ$\;$}}}
\newcommand{\s}{{Succ\;}}

\begin{document}
\homework{7}{18 September}

\section{}
\problem{Read chapter 3 of Bird.}  \problem{Do exercises 3.2.1
\footnote{{\em{i.e.}} prove $\forall m: Nat. \; \s Zero \times m = m$} and
3.2.4 (for finite natural numbers) using the principle of mathematical
induction given on page 63. Hint for 3.2.4: Choose arbitrary $m,n\in{}Nat$ and
do induction on $p$.  }



\section{Induction Examples from Class}

In class we gave the following definitions.

\begin{program*}
\> date Nat = Zero | \sS Nat   deriving (Eq,Ord,Show) \\
\> \\
\> (+) :: Nat -> Nat -> Nat \\
\> m + Zero = m \\
\> m + (\sS n) = \sS (m + n)\\
\>
\end{program*}

We can prove properties of our addition function, verifying that it really
behaves like the addition we know an love, by using mathematical induction.


\begin{lemma}
\label{lemma1}
$\forall{}m:Nat. \;\; Zero + m = m $

\Proof
By induction on $m$.  The property $P$ is given as:
\[P(m) \definedAs Zero + m = m \]
\CASE[$P(Zero)$] We must show $Zero + Zero = Zero$, which holds by the
definition of $+$ so the base case holds.\\

\CASE[$P(\s k)$] Assume $P(k)$ and show $P(\s k)$.
\[\begin{array}{rl}
P(k) :& Zero + k = k \\
P(\s k):&  Zero + (\s k) = (\s k) 
\end{array}\]

Starting on the left side of the equality  $P(\s k)$:

\[Zero + (\s k) \byDef{(+)} \s(Zero + k) \byEq{P(k)} \s k\]

\qed

\end{lemma}

\newpage

\begin{lemma}
\label{lemma2}
$\forall{}n:Nat.\;\forall{}k:Nat.\; \s k + n = k + \s n$
\Proof
By induction on $n$. Then \[P(n)\definedAs \forall{}k:Nat.\; \s k + n = k + \s n \]

\CASE[$P(Zero)$] We must show $\forall{}k:Nat.\; \s k + Zero = k + \s
Zero $. Choose arbitrary $k$ and show $\s k + Zero = k + \s Zero$.  But, consider the following sequences of equalities:
\[\begin{array}{l}
\s k + Zero \byDef{(+)} \s k \\
k + \s Zero \byDef{(+)} \s (k+ Zero) \byDef{(+)} \s k
\end{array}\]
so the base case holds.

\CASE[$P(\s m)$] Assume $P(m)$ and show $P(\s m)$.
\[\begin{array}{rl}
P(m) :& \forall{}k:Nat.\; \s k + m = k + \s m \\
P(\s m):&  \forall{}k:Nat.\; \s k + \s m = k + \s (\s m) 
\end{array}\]

Notice that in the second equation, substituting $\s m$ for $n$ in the term $\s
n$ on the right side gives $\s (\s m)$, this could would be an easy place to
make an error. To prove $p(\s m)$ holds, choose arbitrary $k\in{}Nat$ and show
$\s k + \s m = k + \s (\s m) $ Consider the following sequences of equalities:
\[\begin{array}{l}
\s k + \s m \byDef{(+)} \s(\s k + m) \byEq{P(m)} \s(k + \s m)\\ 
k + \s (\s m)  \byDef{(+)} \s(k + \s m)
\end{array}\]



\qed
\end{lemma}

Now, we prove the commutativity of addition:

\begin{theorem} $\forall{}n: Nat.\; \forall {} m: Nat .\; m + n = n + m $
\Proof By induction on $m$. Then \[P(n)\definedAs \forall {} m: Nat .\; m + n = n + m \] 

\CASE[$P(Zero)$] We must show $\forall {} m: Nat .\; m + Zero = Zero + m
$. Choose arbitrary $m$ and notice that by definition of (+) $m + Zero = m$ and
by Lemma~\ref{lemma1} $Zero + m = m$.  Thus, the base case holds.



\CASE[$P(\s m)$] Assume $P(m)$ and show $P(\s m)$.
\[\begin{array}{rl}
P(k) :&  \forall {} m: Nat .\; m + k = k + m \\
P(\s k):&  \forall {} m: Nat .\; m + \s k = \s k + m  
\end{array}\]

To show $P(\s k)$ choose arbitrary $m$ and show $m + \s k = \s k + m$.
\[\begin{array}{l}
m + \s k \\
\byDef{(+)} \s (m + k)  \\
\byEq{P(k)} \s(k + m) \\
\byDef{(+)[backwards]} k + \s m  \\
\byEq{(Lemma~{\ref{lemma1}})} \s k + m\\
\end{array}\]
Note: In the second to last step, we use the equality $ k + \s m \byDef{(+)}
\s(k + m)$ in the right to left direction. This can be seen as a step in which
we {\em{fold up}} the definition of $(+)$.

\qed
\end{theorem}





\end{document}
% Local Variables:
% mode:latex
% comment-column:0
% comment-start: "% "
% compile-command: "pdflatex hw7a"
% fill-column:79
% End:




