\documentclass[11pt]{article}

\include{header}

\begin{document}
\homework{9}{28 September}

\section{Disjoint Unions}

\exercise{Reread section 2.5 of Bird.}

The Hugs implementation of the the type {\tt{Either}} differs from the
older version of Haskell described in the book in that the following
code is in the prelude.

\begin{program*}
\> data  Either a b  =  Left a | Right b   deriving (Eq, Ord, Read, Show) \\
\>  \\
\> either               :: (a -> c) -> (b -> c) -> Either a b -> c \\
\> either f g (Left x)  =  f x \\
\> either f g (Right y) =  g y 
\end{program*}

So, in Hugs, we use the destructor {\tt{either}} rather than {\tt{case}} as described in Bird. In class, we said:

\begin{program*}
\> case                :: ((a -> c), (b -> c)) -> (Either a b) -> c \\
\> case (f,g) (Left x)  =  f x \\
\> case (f,g) (Right y) =  g y 
\end{program*}

So, {\tt{either}} is just like {\tt{case}}, except it is  fully curried.

\section{The Unit Type}

There is a type of tuples of length zero written {\tt{()}}.  There are
two inhabitants of theis type, $\bot$ (which inhabits every type) and
the element {\tt{()}}.  This can be a bit confusing since, for the
unit type, the element of the unit type has the same syntax as the
type.

Now, we can make a constant into a function by making it a function of
the unit type.
\begin{program*}
\> pifun :: () -> Float \\
\> pifun () = 3.14159
\end{program*}
In the signature for {\tt{pifun}}, ``{\tt{()}}'' denotes the unit type.  In the
definition of {\tt{pifun}}, ``{\tt{()}}'' denotes the single element of the
unit type.

\newpage
\section{Another encoding of Booleans using Disjoint Unions}

Now, conside the following module.

\begin{program*}
\> module Boolean where  \\
\>  \\
\>   type Boolean = Either () () \\
\>  \\
\>   true, false :: Boolean \\
\>   true = Left () \\
\>   false = Right () \\
\>   \\
\>   ifthenelse :: Boolean -> a -> a  -> a \\
\>   ifthenelse b e1 e2 = either ($\backslash$() -> e1) ($\backslash$() -> e2) b 
\end{program*}
Here, the type {\tt{Boolean}} is defined as a disjoint union of the single
element unit type.  So, there are four elements, 

\begin{program*}
\> Boolean = \{Left(), Right(), $\bot$, Left $\bot$, Right $\bot$\} 
\end{program*}
The proper values of the type are {\tt{Left ()}} and {\tt{Right ()}}.

Now, we define constants {\tt{true}} and {\tt{false}} to be {\tt{Left ()}} and
{\tt{Right ()}} respectively.

The definition of {\tt{ifthenelse}} is somewhat interesting.  The function
{\tt{either}} is expecting functions as its first and seconf arguments. We want
{\tt{ifthenelse}} to just evaluate to the expression {\tt{e1}} if the condition
{\tt{b == true}} (recall {\tt{true}} is defined to be {\tt{Left()}}) and to
evaluate to {\tt{e2}} if the conditon {\tt{b==false}} (where {\tt{false}} has
been defined as {\tt{Right()}}).  We wrap {\tt{e1}} and {\tt{e2}} up as lambda
expressions whose argument is the constant {\tt{()}} {]em{i.e.}} the single
element of unit.  This may seem odd -- but the interpreter is using pattern
matching -- and since the only that that can be unwrapped by {\em{either}} if
the argument {\tt{b}} is {\tt{true}} or {\tt{false}} is {\tt{()}}, these
functions {\tt{($\backslash$() -> e1)}} and {\tt{($\backslash$() -> e2)}} will
only ever be applied to the constant {\tt{()}}.\footnote{As an experiment, you
might try typing in the following expressions to the interpreter:
{\tt{($\backslash$1 -> 2)1}} and \goodbreak {\tt{($\backslash$3 -> 2)1}} to see what
Haskell does.}


\exercise{Using the definition of {\tt{ifthenelse, true}} and {\tt{false}} given, prove that 
\begin{program*}
\> ifthenelse true e1 e2 = e1 \\
\> ifthenelse false e1 e2 = e2 
\end{program*}
}


\exercise{Using definitions of {\tt{ifthenelse, true}} and {\tt{false}} given
above, extend the code to include definitions of {\tt{and}}
({\tt{/$\backslash$}}), or ({\tt{$\backslash$/}}),  and implies
({\tt{.=>}}). 
\begin{program*}
\>   (/$\backslash$),($\backslash$/),(.=>) :: Boolean -> Boolean -> Boolean
\end{program*}
You may use the destructor {\tt{either}} directly if you like, or define the
functions in terms of {\tt{ifthenelse}}.  There is a link on the hw page to
{\tt{hw9\_expected.txt}} which contains an example test run of your code showing
what the results should look like.}


\end{document}
% Local Variables:
% mode:latex
% comment-column:0
% comment-start: "% "
% compile-command: "latex hw9; dvips hw9 -o hw9.ps; pdflatex hw9"
% fill-column:79
% End:




