\documentclass[11pt]{article}

\include{header}

\begin{document}
\homework{9}{27 September}

\section{List Comprehensions}

As mentioned in class, list comprehensions provide a syntactically
concise and expressive means of writing list functions and every
Haskell programmer should know how to use them.  You should read about
them in Bird (pp. 112-115) and in LYAHFGG (Chapter 2).

As examples (in addition to the ones in the readings) {\it{take}} and {\it{drop}} can be implemented using list comprehensions as follows:

\begin{program**}
\> take' k xs = [ xs !! i $\mid$ i $\leftarrow$ [0..(k-1)]] \\
\> drop' k xs = [ xs !! j $\mid$ j $\leftarrow$ [k..(length xs - 1)]]
\end{program**}

Note that {\it{!!}} is the select operator for indexing into a list, thus
{\it{(xs !! j)}} returns the $(j+1)^{st}$ element (zero based indexing ) of the
list $xs$. {\it{[0,1,2] !! 1 == 1}} (the index is 1 but 1 is the second element
in the list.)


\problem{}

In this assignment you will use list comprehensions to implement the
functions to support operations on a toy relational database
representing the personnel files of the ACME Programming Company.
Tables are represented in Haskell as lists of lists.
\ \\
We define some datatypes to represent the database.

\begin{program**}
\> data Position = CEO $\mid$ Manager $\mid$ Programmer $\mid$ Intern  deriving (Eq,Show)\\
\> \\
\> data Field = EmployeeId Int $\mid$ T Position $\mid$ Name String $\mid$ Salary Int deriving (Eq) \vspace{1em}\\
\> instance Show Field where  \\
\>    show (EmployeeId k) = show k  \\
\>    show (T p) = show p  \\
\>    show (Name s) = s  \\
\>    show (Salary k) = show k  \\
\> \\
\> type Column = Int \\
\> type Row = [Field]\\
\> type Table = [Row]\\
\end{program**}


\noindent{}Here are some tables.\vspace{-1em}

\begin{program**}
 \=\=                                                         \= \=   \kill\\
\> employees =                    \>        \> positions =   \\			    
\>  [[EmployeeId 1234, Name "Smith"],       \>\>\> [[EmployeeId 1234, T CEO],  \\	    
\>   [EmployeeId 1235, Name "Jones"],       \>\>\> [EmployeeId 1235, T Manager],   \\	    
\>   [EmployeeId 1236, Name "Brown"],       \>\>\> [EmployeeId 1236, T Programmer],   \\  
\>   [EmployeeId 1237, Name "Green"],       \>\>\> [EmployeeId 1237, T Programmer],  \\   
\>   [EmployeeId 1238, Name "Edwards"]]     \>\>\> [EmployeeId 1238, T Intern]]  \\       
\> salary =                                 \>\> stock\_options =   \\		 
\>  [[EmployeeId 1234, Salary 1000000],     \>\>\> [[Id 1234, Shares 1000000],  \\	 
\>   [EmployeeId 1235, Salary 100000],      \>\>\> [Id 1235, Shares 1000],   \\	 
\>   [EmployeeId 1236, Salary 55000 ],      \>\>\> [Id 1236, Shares 5000 ],   \\	 
\>   [EmployeeId 1237, Salary 78000],       \>\>\> [Id 1237, Shares 2500],  \\	 
\>   [EmployeeId 1238,Salary 0]]            \>\>\> [Id 1238, Shares 5]]  \\         
\>   \\
\end{program**}

\noindent{}Using list comprehensions, you need to implement functions having the following types.

\begin{description} 
\item{{\it{delete :: Column $\rightarrow$ Row $\rightarrow$ Row}}} \\The delete
function removes the specified column from the row.  Remember to implement the
function using a list comprehension.

\item{{\it{join :: (Table,Column) $\rightarrow$ (Table,Column) $\rightarrow$
Table}}} \\ The join function performs a kind of consistent merge on two tables
by matching values in the named columns.  If the call is of the form {\it{join
(t1,i) (t2,j)}} the resulting table is created by appending rows from $t1$ to
rows of $t2$ (from which column j has been deleted) whenever the value in
column $i$ of a row from $t1$ is equal to the value in column $j$ of a row of
$t2$. Deleting column $j$ from rows of $t2$ (before appending them with the
matching rows of $t1$) guarantees that the join operation does not duplicate a
copy of the column.  Examining the expected output for the test cases should
make the more clear. Remember, use a list comprehension to implement the
function.

\item{{\it{project :: [Column] $\rightarrow$ Table $\rightarrow$ Table }}} \\In
a call of the form {\it{project columns t}} a new table is constructed from
table $t$ containing the columns of $t$ as listed in the argument
{\it{columns}}.  If the list of columns is empty, the resulting table is
empty. Using a projection it is possible to change the order and number of
occurrences of the columns in a table.  For example, if a $t$ table has three
columns, the projection {\it{project [2,1,0,1] t}} results in a table with four
columns where the first is the third column from $t$, the second is the second
column from $t$, the third is the first column from $t$ and the fourth is
another copy of the second column from $t$. Remember, you need to use a list
comprehension to implement the function.
\end{description}




\end{document}
% Local Variables:
% mode:latex
% comment-column:0
% comment-start: "\\> "
% comment-end: "\\\\"
% compile-command: "pdflatex hw9c"
% fill-column:79
% End:




